A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2019-01-27 article/04_Discussion.Rmd
language in discussion

The A sections of the figures \ref{development_burial_type} and \ref{development_burial_construction} with absolute grave numbers per region and year indicate areas and periods with a lot and others with very few points of data. For some there is not enough information to make any decision about the relative dominance and for many the general relevance and accuracy of the resulting time series can be questioned. There's a variety of reasons for this which should be considered for future data acquisition in Radon-B and beyond:

1. The data in Radon-B depends on the projects and interests of the involved researchers: The data was collected as needed for individual regions rather than systematically for all of Europe.

@@

3. In the field of investigation of this work different research traditions meet, which in the past considered ^14^C data more or less essential for the construction of an absolute chronological assessment. In various Metal Age contexts, 14C data are inferior to the accuracy of relative chronological classifications - e.g. by primer typology - and are therefore used only sporadically to create absolute chronological fix points. Dendrochronological dating is preferred in moist soil contexts and when the necessary woods are available. The available ^14^C calibration curves form plateaus in some periods, which significantly limit the possible dating accuracy of the data. If such an effect is known, then no data is ordered at all.

4. ^14^C dating requires a small but sufficiently contextualised amount of organic finds. Due to the old wood and old water effect, short-lived sample materials such as bones of terrestrial organisms or charred cereal grains are preferred. The majority of ^14^C data for this work comes from bones and teeth. However, their preservation is subject to huge differences depending on the soil [@brockPrescreeningTechniquesIdentification2010; @gordonSoilPHBone1981; @nielsen-marshBoneDiagenesisEuropean2007]. In many mineral soil contexts no sufficiently good, dateable material is available. This situation can affect entire regions, if, for example, poor bone preservation prevails in general as a result of low-calcification soils.

5. From very few regions and periods of the European Bronze Age so many graves are documented that it can be assumed that a significant sample of the burial culture of the entire population has been researched. Instead, it is evident in several contexts that only the burials of individual social groups, gender or ethnic group were recorded: The quantity and nature of the known graves cannot be representative of all the deceased. In this case, it can be assumed that further, deviant burial rituals were practised, which left no remains or remains that were difficult to grasp archaeologically. These rituals are not shown in the ^14^C data from Radon-B.

Concerning the primary variables *burial type* and *burial construction*, further special effects must be taken into account: The disturbance and robbery of graves was a widespread phenomenon in history and prehistory. Burial mounds were usually attacked more severely than flat graves due to their better visibility and could, therefore, be underrepresented in Radon-B. In many places, they were also the subject of archaeological expeditions in the early days of archaeology at the end of the 19th and beginning of the 20th century. Of course, these excavations did not yield ^14^C-dates. Irrespective of this, it can be assumed that a large part of the burial mounds erected in the Bronze Age was destroyed by agricultural activity. Ploughing could also result in an under-representation of the archaeological findings [@hardingEuropeanSocietiesBronze2000a]. On the other hand, if a burial mound is still present, it is more readily visible and recognised by archaeologists today: Mounds are often clearly identifiable in aerial photographs or digital terrain models, while flat graves are much more difficult to identify for today's remote sensing methods. It is difficult to estimate the impact of these effects in detail, but an overrepresentation of mounds seems more likely than an overrepresentation of flat graves. The dating of cremation graves on bone remains methodically differs from the dating of body graves because it is based on the extraction of the mineral, not the organic, fraction in the bone [@lantingDatingCrematedBones2001]. The underlying chemical method was developed later, has only been used more frequently in the past 15 years and is associated with some particular difficulties [@olsenCharacterisationBlindTesting2008; @strydonck14CDatingCremated2009]. Burnt bone is generally more susceptible to weathering than unburned bone. An under-representation of cremation graves in current data collections is, therefore, more plausible.

>>>>

The **A** sections of the figures \ref{development_burial_type} and \ref{development_burial_construction} with absolute grave numbers per region and year indicate areas and periods with a lot and others with few points of data. For some there is not enough information to make any decision about the relative dominance and for many the general relevance and accuracy of the resulting time series can be questioned. There is a variety of reasons for this. These should also be considered for future data acquisition in Radon-B and beyond.

1. The data in Radon-B depend on the projects and interests of the participating researchers: It was collected as needed for individual regions rather than systematically for all of Europe.

@@

3. Different research traditions in archaeology considered ^14^C data in the past more or less essential for the reconstruction of absolute chronology. In various Metal Age contexts, ^14^C data provide inferior precision in comparison to relative chronological classifications - e.g. by fibula typology - and are therefore used only sporadically to create absolute chronological fix points. Dendrochronological dating is preferred in waterlogged deposits and whenever the necessary woods are available. ^14^C calibration curves form plateaus in some periods, which significantly limit the possible dating precision. If such an effect is known, then often no dates are commissioned at all.

4. ^14^C dating requires a small but sufficiently contextualised amount of organic finds. Due to the old wood and old water effect, short-lived sample materials such as bones of terrestrial organisms or charred cereal grains are preferred. The majority of ^14^C data for this work comes from bones and teeth. However, their preservation is subject to huge differences depending on the soil [@brockPrescreeningTechniquesIdentification2010; @gordonSoilPHBone1981; @nielsen-marshBoneDiagenesisEuropean2007]. In many mineral soil contexts no sufficiently good, dateable material is available. This situation can affect entire regions, if, for example, poor bone preservation prevails as a result of lime-deficient soils.

5. In very few regions and periods of the European Bronze Age the number of excavated graves is sufficiently big to assume that a statistically relevant sample of the burial culture of the entire population is documented. Instead, it is evident in most contexts that only the burials of certain social groups, genders or ethnic groups were recorded. The quantity and nature of the known graves cannot be representative. In this case it can be expected that further, deviant burial rituals were practised, which left no archaeologically visible remains. These rituals are invisible in the ^14^C data from Radon-B.

Concerning the primary variables *burial type* and *burial construction*, further special effects must be taken into account: The disturbance and robbery of graves was a widespread phenomenon in history and prehistory. Burial mounds were usually attacked more severely than flat graves due to their better visibility and could, therefore, be under-represented in Radon-B. In many places, they were also the subject of expeditions in the early days of archaeology at the end of the 19th and beginning of the 20th century. Of course, these excavations did not yield ^14^C-dates. Irrespective of this, it can be assumed that a large part of the burial mounds erected in the Bronze Age was destroyed by agricultural activity. Ploughing could also result in an under-representation of the archaeological findings [@hardingEuropeanSocietiesBronze2000a]. On the other hand, if a burial mound is still present, it is more readily visible and recognised by archaeologists today: Mounds are often clearly identifiable in aerial photographs or digital terrain models, while flat graves are much more difficult to identify for today's remote sensing methods. It is difficult to estimate the impact of these effects in detail, but an over-representation of mounds seems more likely than an over-representation of flat graves. The dating of cremation graves on bone remains methodically differs from the dating of inhumations because it is based on the extraction of the mineral, not the organic, fraction in the bone [@lantingDatingCrematedBones2001]. The underlying chemical method was developed later, has only been used more frequently in the past 15 years and is associated with some particular difficulties [@olsenCharacterisationBlindTesting2008; @strydonck14CDatingCremated2009]. Burnt bone is generally more susceptible to weathering than unburned bone. An under-representation of cremation graves in current data collections is, therefore, more plausible.
